---
title: "1.0.0 - Beginner - Introduction to Statistical Concepts (325 Prep Course)"
author: COMET Team <br> _Charlotte White, Anneke Dresselhuis_, _Jonathan Graves_
date: 10 August 2023
description: "This notebook introduces you to some fundamental statistics and basic probability concepts. It is designed to cover introductory statistics material taught in Foundations of Mathematics 12 and ECON 226: Making Sense of Economic Data for students who either haven’t taken these courses or who would like a refresher before going into ECON 325."
categories: [introduction, beginner, probability, data, descriptive statistics, data visualization, inferential statistics, probability notation, joint-probability, distributions, data types, R, econ 325]
toc: true
execute:
  eval: false
format: 
  html: default
  ipynb:
    jupyter:
      kernelspec:
        display_name: R
        language: r
        name: ir
---

## Outline

### Prerequisites:

This notebook uses the [Jupyter system](https://jupyter.org/), which is
a connected system of open-source tools designed to support interactive,
web-based computing. This system allows you to run cells of code, answer
questions and apply your learning as you go. To familiarize yourself
with the Jupyter environment, please check out the following resources
before continuing with this lesson.

### Introduction to Jupyter

Additional resources for familiarizing yourself with the Jupyter:

-   [Using COMET as a
    student](https://comet.arts.ubc.ca/pages/installing_locally.html)

-   [Quickstart Guide](https://comet.arts.ubc.ca/pages/quickstart.html)

### Learning Objectives

-   Describe the probability of an event, using accurate terms
    -   Understand how to read basic probability notations
    -   Calculate the theoretical and experimental probability
-   Describe and apply the fundamental rules of counting
-   Gain familiarity with data types and their functions
-   Identify common data visualizations and their uses
-   Interpret regression equations and apply a regression to real-world
    examples
-   Describe the connection between probability and statistical
    inferences

### References

-   \[Appleby & Ranieri - iWrite Math 12: Foundations of Math, British
    Columbia Edition\]
-   [Timbers, Campbell, and Lee - Chapter 10: Statistical
    Inference](https://datasciencebook.ca/inference.html)
-   \[De Veaux, Velleman, and Bock - Intro Stats\]
-   [Illowsky and Dean - Chapter 12: Linear Regression and
    Correlation](https://openstax.org/books/introductory-statistics/pages/12-introduction)

Before beginning the lesson, please run the below cell.

```{r}
#| output: false
source("beginner_intro_to_statistical_concepts_functions.r")
source("beginner_intro_to_statistical_concepts_tests.r")
```

## PART 1: PROBABILITY

Probability is an area of study that deals with the mathematics of
chance or prediction. This lesson is designed to prime you with some
basic, introductory concepts that are essential for understanding both
statistics and probability.

**Material Covered:**

-   Random events

-   Formal probability (terminology and notation)

-   Calculating theoretical and experimental probability

-   Unions, intersections, conditional probability and independence

-   Models for probability

To begin, let's familiarize ourselves with some of the basic terms used
in the study of probability.

-   **Random**: an situation is random if we know the possible results
    it can have, but not which particular result has or will occur.
-   **Trial**: a single occasion where we observe a random situation.
    Since the situation is random, the result of the trial cannot be
    predicted with certainty
-   **Outcome**: the result measured or reported from the trial
-   **Sample space**: all possible outcomes for a trial
-   **Event**: a group of one or more outcomes in a sample space
-   **Probability notation**: a formal notation method of efficiently
    conveying probability concepts

Although these may seem like abstract concepts, the reality is that we
encounter these concepts in our lives on a daily basis.

Think about rolling a standard, six-sided die.

![A standard six-sided die, used for a number of games (Credit: [Roger
McLassus](https://commons.wikimedia.org/wiki/File:Dice_2005.jpg))](media/Dice_2005.jpg){fig-alt="game die"}

-   Rolling the die would be considered the *trial* in this example,
    which could result in six distinct *outcomes* based on which side of
    the die is facing up: 1,2,3,4,5,6.
-   To help differentiate the outcomes from other mathematical objects,
    we can use *probability notation* and list all possible outcomes
    within curly brackets, separated by commas: {1,2,3,4,5,6}.
-   This list of outcomes makes up the *sample space*.
-   We could also group our outcomes into *events*, such as 'rolling an
    even number' or 'rolling an odd number'.

We can think about probability in terms of *likelihood*, which means how
"likely" an event is. If we are familiar with the circumstances
surrounding an event (the trial, outcome(s), and sample space) we are
able to understand the likelihood of an event. For each event, there is
both a theoretical and experimental probability[^1]. The theoretical
probability can be calculated as follows:

[^1]: Sometimes also called the *empirical* probability

$$ 
P(A) = \frac{\text{\# of outcomes in A }}{\text{\# total possible outcomes in the sample space }}
$$

Where $P(A)$ is the probability notation which describes the
**probability** of event $A$ occurring.

When you toss a die, there is a theoretical probability for each event
possible, which depends on the outcomes which make up the event.

E.g $P(\text{rolling an odd number})$= number of outcomes of rolling an
odd number, which is three (rolling a one, a three, or a five), divided
by the total possible outcomes which we know to be six. Therefore,
$P(\text{rolling an odd number}) = 0.5$.

$$ 
P(Odd) = \frac{\text{# outcomes of rolling an odd number (1,3,5)}}{\text{# total possible outcomes}}=\frac36=0.5
$$

By these kinds of calculations, we can assign a numerical value between
0 and 1 to the likelihood of an event occurring.

So, if you roll a die ten times, are you guaranteed to roll five even
numbers and five odd numbers? Not necessarily; however, if you were to
conduct these trials and record your results you could calculate the
experimental probability.

$$ 
P(A) = \frac{\text{# of times event A occurs}}{\text{# total number of trials}}
$$

Experimental probability is calculated on the basis of events that
occurred, whereas theoretical probability is based purely on
mathematical reasoning.

With a small number of trials, there may be large differences in these
two values. It's conceivable that if you rolled a die four times, you
could get an even number every time, making
$P(\text{rolling an odd number}) = 0$. That's very different from the
value of $0.5$ that we calculated earlier.

What if you rolled a thousand times, though? Intuitively, you might
already have realized that the more times we repeat this random process,
the more likely it is that the experimental probability of our event of
interest will settle towards its expected probability.

**Question 1:** True or False? The combined probability of all outcomes
in a sample space will always be equal to 1. Try it with the dice
example!

```{r}
#| tags: [question]
#| eval: false

answer_1 <- '' # your answer here('true' or 'false')

test_1()
```

```{r}
#| tags: [answer]
#| eval: false

answer_1 <- '' # your answer here('true' or 'false')

test_1()
```

So we now know how to calculate the probability that an event will
occur. What about the probability that it will *not* occur? This is
called the **complementary event**.

In a situation where there are only two possible outcomes, the
complementary event is relatively straightforward. Consider a coin toss;
what is the complement of the event "lands on heads"? If you're thinking
"lands on tails", you would be partially right. The complementary event
would really be "does not land on heads". Since there's only two
outcomes in the sample space, then "lands on tails" is the only way to
achieve the complementary event in this example.

However, this changes when there are more possible outcomes -- like
rolling a six-sided die. Now, the probability of the complement to
"rolling a one" is the combined probability of rolling a two, three,
four, five, or six.

### Case Study

To better understand these ideas, we will use a case study: *salmon
fishing* in British Columbia.

In BC, salmon are an important economic resource both for commercial
sales and personal consumption. They also are a source of livelihood and
with high cultural significance for some communities. They play a major
role as a keystone species in coastal ecosystems for their critical
place in the food chain and significant contributions to nutrient
cycling throughout their life cycles, which keep the surrounding
streams, trees, and soils healthy. Given their importance, there is a
need to maintain strong records of populations to inform decisions about
fishing regulations and habitat protections in order to maintain strong
salmon populations.

![A sockeye salmon, one of BC's most distinctive fish (Credit: [Milton
Love](https://commons.wikimedia.org/wiki/File:Sockeye_salmon_swimming_right.jpg))](media/Sockeye_salmon_swimming_right.jpg){fig-alt="sockeye salmon"}

![Chinese immigrant Joe Sum fishing for Salmon in Vancouver, BC, c.
August 1945. Many immigrants and Indigenous people worked in salmon
fisheries and canneries during the 20th century. (Credit: [Library and
Archives
Canada](https://recherche-collection-search.bac-lac.gc.ca/eng/home/record?app=fonandcol&IdNumber=3601932))](media/Joe_Sum_salmon_fishing.jpg){fig-alt="Chinese immigrant Joe Sum fishing salmon"}

Imagine you're out fishing for salmon. Each time you cast out your line,
there are five possible types of salmon it may come up with: *Sockeye*,
*Chum*, *Coho*, *Chinook*, and *Pink*. For the following questions,
assume that each type of salmon is equally likely to bite your fishing
rod.

**Question 2:** What is the theoretical probability of catching a Coho
salmon? Please give your answer to two decimal places.

```{r}
#| tags: [question]
#| eval: false

answer_2 <- # your answer here

test_2()
```

```{r}
#| tags: [answer]
#| eval: false

answer_2 <- 0.2 # your answer here

test_2 
```

Despite the integral role of salmon in supporting BC's coastal
ecosystems and communities, data show that many salmon populations are
declining due to pressures from climate change, habitat loss and
pollution, and fishing. As part of its response, the department of
Fisheries and Oceans Canada restricts the [type and size of
fish](https://www.pac.dfo-mpo.gc.ca/fm-gp/rec/bc-zones-cb-eng.html#management-areas)
commercial fishers are allowed to catch and retain. Off the coast of
Salt Spring Island, for example, it is not permitted to catch and take
home any Coho, Chinook, or Sockeye salmon.

**Question 3:** Suppose you are fishing off Saltspring Island. With this
regulation in effect, what is the theoretical probability of catching a
salmon that you can take home with you?

```{r}
#| tags: [question]
#| eval: false

answer_3 <- # your answer here

test_3()
```

```{r}
#| tags: [answer]
#| eval: false

answer_3 <- 0.4 # your answer here

test_3()
```

**Question 4:** What is the theoretical probability of the complementary
event? Please give your answer to two decimal places.

```{r}
#| tags: [question]
#| eval: false

answer_4 <- # your answer here

test_4()
```

```{r}
#| tags: [answer]
#| eval: false

answer_4 <- 0.6 # your answer here

test_4()
```

### Events, Relationships, and Probability Notation

We'll consider catching a salmon we can take home event $A$ (Chum,
Pink), and catching one that we cannot take home (Chinook, Sockeye, or
Coho) event $B$. We would denote their probabilities as $P(A)$ and
$P(B)$ respectively. We could also write them as $P(A)$ and $P(A')$,
since these are complementary events.

In this case, $A$ and $B$ are **mutually exclusive events**, meaning
that they have no common outcomes. In any of the possible outcomes of
what salmon you catch, either event $A$ *or* event $B$ will occur, but
never both $A$ and $B$ at the same time. This is the case with all sets
of events which are complementary events, as there are no common
outcomes.

![A chart displaying the outcomes included in events A and
B.](media/diagrams/events_outcomes1.png){fig-alt="A chart displaying the outcomes included in event A (Chum, Pink) and event B(Chinook, Sockeye, Coho)"}

Suppose you're allowed to take home Chum and Pink salmon, but you only
enjoy eating Sockeye, Coho, and Pink salmon. We'll consider catching a
salmon you are allowed to take home as event $A$, and catching a salmon
you *want* to take home event $B$. These events are not mutually
exclusive. They each have an individual probability of occurring, as
well as a joint probability of both events occurring.

![A chart displaying the outcomes included in events A and
B.](media/diagrams/events_outcomes2.png){fig-alt="A chart displaying the outcomes included in event A (Chum , Pink) and event B(Sockeye, Coho, Pink)"}

We can use probability notation when exploring the joint-probability of
two events ($A$,$B$). In probability notation, the symbol $\cup$ is used
to express an **intersection** as follows: $$
P(A\cap B)
$$

This intersection is the joint-probability, or the probability of both
event $A$ *and* $B$ occurring. The event $(A\cap B)$ occurs only when
both $A$ and $B$ occur simultaneously. There is only an intersection
when the events are not mutually exclusive.

![A venn diagram to help visualize the concept of
joint-probability.](media/diagrams/joint_probability.png){fig-alt="A venn diagram of circles A and B, which shows that the joint-probability of events A and B can be visualized as the area of overlap."}

To calculate the probability of the intersection, we need to calculate
and combine the probability of all outcomes that are included in both
events.

**Question 5:** If catching a Pink or Chum salmon is event $A$, and
catching a Pink, Sockeye, or Coho salmon is event $B$, what is
$P(A\cap B)$? Assume that your sample space includes all five species of
salmon, and that each is equally likely to be caught. Please give your
answer to one decimal place.

```{r}
#| tags: [question]
#| eval: false

answer_5 <-  # your answer here

test_5()
```

```{r}
#| tags: [answer]
#| eval: false

answer_5 <- 0.2 # your answer here

test_5()
```

We can also use a similar notation to express a **union**: $$
P(A\cup B)
$$

Event $(A \cup B)$ occurs if either $A$ *or* $B$ *or* both occur. This
can happen regardless of whether $A$ and $B$ are mutually exclusive.

![A pair of venn diagrams to help visualize types of
unions.](media/diagrams/types_of_unions.png){fig-alt="A pair of venn diagrams representing relationships between events A and B; mutually exclusive events have no area of overlap, whereas non-mutually exclusive events do"}

In calculating the probability of a union, however, it does matter if
the events are mutually exclusive. If $A$ and $B$ are mutually
exclusive, then: $$
P(A\cup B) = P(A) + P(B)
$$

If $A$ and $B$ are not mutually exclusive, then:

$$
P(A\cup B) = P(A) + P(B) - P(A\cap B)
$$

The joint probability is subtracted to avoid "double-counting" the area
of overlap between $A$ and $B$.

![A pair of venn diagrams to help visualize probability
calculations.](media/diagrams/calculating_unions.png){fig-alt="A pair of venn diagrams visualizing the calculation of the probability for union."}

**Question 6:** If catching a Pink or Chum salmon is event $A$, and
catching a Pink, Sockeye, or Coho salmon is event $B$, what is
$P(A\cup B)$? Assume that your sample space includes all five species of
salmon, and that each is equally likely to be caught. Please give your
answer to one decimal place.

```{r}
#| tags: [question]
#| eval: false

answer_6 <-  # your answer here

test_6()
```

```{r}
#| tags: [answer]
#| eval: false

answer_6 <- 0.8 # your answer here

test_6()
```

### Conditional Probability and Independence

If you toss a coin and it lands on heads, the probabilities of the
outcomes for your next coin flip remain unchanged. This is because
tossing a coin is an **independent event**. Even though we are expecting
an approximately equal number of heads and tails outcomes from the total
number of tosses we perform, we know that the outcome of our toss has no
power to dictate the following coin toss.

Let's return to our fishing example. Say you cast out your line once and
catch a Chinook salmon, put it in your boat, then you cast out your line
again. Has the probability of your outcome changed because you caught a
Chinook salmon? What about if you had put the Chinook back in the water
after you'd caught it?

It might be challenging to conceptualize how these events would impact
each other in such a large body of water like the ocean. Think about how
this would work if instead we had a pond, with only a few salmon.

![A diagram of fish in a pond, showing how probability changes as the
number of a certain type of fish
decreases](media/diagrams/fishing_pond.png){fig-alt="A visual of a pond with ten salmon, and the resulting pond of nine salmon after one salmon as been removed."}

You can easily see now how removing a fish from the pond after you catch
it would affect the probability of the next fish you catch. In this
case, the succession of trials would *not* be independent.

**Conditional probability** allows us to understand the probability of
events as they relate to the probability of other events.

See the below contingency table showing some sample data that could have
been collected about salmon caught in a certain region.

![A contingency table showing a possible configuration of hatchery and
wild salmon caught for two different
species](media/diagrams/contingency_table.png){fig-alt="A contingency table; wild Coho = 32, hatchery Coho = 60, total Coho = 92, wild Sockeye = 15, hatchery Sockeye = 50, total Sockeye = 65, total salmon = 157"}

Take a minute to familiarize yourself with the table. In total there
were 157 salmon caught in the sampling period. One hundred and ten of
them were hatchery salmon, forty-seven were wild. Sixty-five of them
were Sockeye, ninety-two of them were Coho, and wild and hatchery fish
were represented for both species in different amounts, as you can see
from the values in the middle of the table.

You are able to easily see that the probability of catching a wild
salmon is $47/157 = 0.299$. Would this probability change if we knew
that we were only going to catch Coho? You might notice that wild salmon
make up a much larger percentage of the Coho, so we would have a higher
probability of catching a wild salmon in the case where we know it will
be a Coho salmon. This is because we're now observing a *conditional*
distribution of the sixty-five Coho salmon. This probability is
expressed as:

$$
P(A|B)
$$

Where $A$ would be the event 'catching a wild salmon' and $B$ would be
the event 'catching a Coho salmon'. $P(A|B)$ then means 'the probability
of catching a wild salmon *given* that it's a Coho salmon'. In this case
$P(A|B) = 32/92 = 0.35$.

**Question f:** What is the probability of catching a sockeye salmon
given that it's a wild salmon? Please round your answer to two decimal
places.

```{r}
#| tags: [question]
#| eval: false

answer_6.1 <- # your answer here

test_6.1()
```

```{r}
#| tags: [answer]
#| eval: false

answer_6.1 <- 0.32 # your answer here

test_6.1()
```

### Distributions

In the previous example, we began exploring instances where the
probabilities are not equal among the different outcomes. In controlled
settings like rolling dice or tossing coins, we can easily see why we
would assume equal probability in events- these specific trials have
been designed to give no advantage to one outcome or another.

However, for other random events, the probability of outcomes is not
equal. When you cast your line into the water, the probability that you
will catch a given species of salmon is not going to be equal, and may
depend on where you are and the abundance of different salmon species in
that area, or any other number of factors. This is a **probability
distribution** which refers to the pattern or arrangement of
probabilities in a population.

Over the course of an afternoon, you cast your line out 15 times. If the
probabilities of catching each of the five salmon species were equal,
your distribution would look like this:

| Salmon  | Frequency | Probability |
|---------|-----------|-------------|
| Pink    | 3         | 1/5         |
| Sockeye | 3         | 1/5         |
| Chum    | 3         | 1/5         |
| Coho    | 3         | 1/5         |
| Chinook | 3         | 1/5         |

But experimentally, that's not what happens. Instead you find somethings
like this:

| Salmon  | Frequency | Probability |
|---------|-----------|-------------|
| Pink    | 2         | 2/15        |
| Sockeye | 1         | 1/15        |
| Chum    | 4         | 4/15        |
| Coho    | 7         | 7/15        |
| Chinook | 0         | 0           |

Note that the probabilities still total 1.

Distributions are usually described as functions: the function takes in
a possible event, then reports back the probability of that event
occurring.

### Probability Models

One helpful tool to help understand this better is the use of
**probability models**, which is a mathematical description of an
experiment. Using probability models, we can use our understanding of
probabilities to predict random behaviour.

The below table displays a sample probability distribution for the
number of fish caught in a day.

| Outcome (number of fish) $x$ | Probability $P(x)$ |
|------------------------------|--------------------|
| 0                            | 0.15               |
| 1                            | 0.23               |
| 2                            | 0.28               |
| 3                            | 0.22               |
| 4                            | 0.12               |

Given these probabilities, how many fish can we expect to catch on a day
out? While there is no way of knowing exactly what will happen on a
given day, we can figure out what to expect by calculating the
**expected value**. The equation to establish this is:

$$
E(X) = \sum_x x \cdot P(x)
$$

Where $E(x)$ is the expected value of random variable $X$, $x$ is a
particular value that the random variable can take on, and $P(x)$ is the
probability of that value occurring. In the following example, we will
use this equation to determine the answer to our previous question: how
many fish can we expect to catch in a day?

$$
E(X) = \sum_x x \cdot P(x)\\
$$ $$
E(X) = (0\cdot 0.15)+(1\cdot 1.23)+(2 \cdot 0.28)+(3 \cdot 0.22)+(4 \cdot 0.12)
$$ $$
E(X) = 1.93
$$

So this means that on average, we can expect 1.93 fish to be caught in
one day. This doesn't mean that we certainly will catch this number of
fish each day. In fact, as you might have noticed, it isn't possible to
catch a fraction of a fish. This is simply an average.

If we're fishing for recreational purposes, it may not be as important
to know this number. Consider how that might change if fishing is a part
of our livelihood. Knowing the number of fish we can expect to catch
might then help us to make decisions about how much time we need to
spend fishing, or how to price our fish. This is a common application of
probability models.

### Exercise

Now it's your turn. The below sample probability distribution shows the
average size of different salmon species in BC.

| Salmon  | Length (cm) | Probability |
|---------|-------------|-------------|
| Pink    | 51          | 0.27        |
| Sockeye | 68          | 0.55        |
| Chum    | 76          | 0.18        |

**Question 6.2:** What is the expected length in cm of a salmon that you
catch? Please give you answer to two decimal places.

```{r}
#| tags: [question]
#| eval: false

answer_6.2 <- # your answer here

test_6.2()
```

```{r}
#| tags: [answer]
#| eval: false

answer_6.2 <- 64.85 # your answer here

test_6.2()
```

## PART 2: COUNTING, PERMUTATIONS, AND COMBINATIONS

You may have noticed that the probability of an outcome had a lot to do
with the *possibilities* that existed for a trial. In determining
probability, we often will need to know that total number of possible
outcomes, and what those outcomes are. To do that, we need to have
efficient ways to count them!

**Material Covered:** \* Fundamental counting principles \* Calculating
permutations with repeated elements \* Calculating combinations

Imagine you're preparing to go on a fishing trip. This could look
different for everyone, but below is a list of options for how you might
approach different aspects of your trip.

| Transportation | Location | Fishing Method |
|----------------|----------|----------------|
| Driving        | Estuary  | Net            |
| Walking        | River    | Weir           |
|                |          | Rod            |

> *Fishing Weirs*: You may or may not be familiar with the term 'weir',
> which is a structure placed across a channel of water to trap fish. In
> BC, many Indigenous fishers used this technology to not only catch
> fish, but to collect information on the size of the year's salmon run,
> which could inform harvesting practices. Weirs were banned by the
> Canadian government in an effort of colonization, but in recent years,
> some coastal communities have been reviving the practice as an
> effective method to continue monitoring salmon populations.

![A fishing weir from the mid-1800s on the Cowichan River (Credit:
[Fredrick
Dally](https://commons.wikimedia.org/wiki/File:Salmon_weir_at_Quamichan_Village_on_the_Cowichan_River,_Vancouver_Island.jpg))](media/Salmon_weir_at_Quamichan_Village.jpg){fig-alt="Fishing Weir"}

How many different ways could we go about this fishing trip if we have
to choose just one method at each level? To illustrate this, we'll draw
out our options as a decision tree.

![A tree diagram of the possible ways to go on a fishing trip based on
the above
table.](media/diagrams/tree_diagram1.png){fig-alt="A tree diagram summarizing all 12 possible outcomes fishing trips from the list of options."}

When it's drawn in a tree format like this, we can simply count up the
total number of outcomes. When we don't want to take the time to draw a
diagram, or the numbers are too large to do so, we can also calculate
the total outcomes using the **fundamental counting principle**. By this
principle, we know that if there are *m* choices at the first stage, and
*n* choices at the second stage, then the total number of outcomes is
$m \cdot n$.

$$
 \text{Total # of ways} = \text{2 transportation} \times \text{2 location} \times \text{3 fishing methods} = 12
$$

We'll continue to illustrate this with the example of salmon. If we were
to catch any five salmon, we can create several different arrangements
based on the species of salmon we have caught.

![A table showing the possible outcomes for each
"spot"](media/diagrams/possible_outcomes.png){fig-alt="A table showing the five possible options of species (Pink, Coho, Sockeye, Chinook, Chum) for each of the five fish caught"}

**Question 7:** Use the fundamental counting principle to calculate how
many arrangements you can make using all five fish. Give your answer as
a whole number, with no decimals. *Hint: the total number of "levels"
you have is five, and each species of fish represents a possibility at
each level.*

```{r}
#| tags: [question]
#| eval: false

answer_7 <- # your answer here

test_7()
```

```{r}
#| tags: [answer]
#| eval: false

answer_7 <- 3125 # your answer here

test_7()
```

Now what if our salmon were in a small pond again, and there is only one
of each species? This means that at each level, we would have one less
possible species to draw than the previous level.

**Question 8:** If we start with five possible species, how many total
arrangements of salmon will we have?

```{r}
#| tags: [question]
#| eval: false

answer_8 <- # your answer here

test_8()
```

```{r}
#| tags: [answer]
#| eval: false

answer_8 <- 120 # your answer here

test_8()
```

In this particular example, we could also write the answer out using
**factorial notation**. Factorial notation expresses a product of all
positive integers which lead up to a number. In the below equation, the
factorial notation is on the left, expressing the full formula on the
right.\
$$
5! = 5 \times 4 \times 3 \times 2 \times 1 
$$

This provides us with a way to simplify equations, and becomes important
as we continue to build an understanding of counting.

In our previous examples, we were using the fundamental counting rule to
calculate the number of arrangements we could generate from a given set
of elements (salmon species). In these examples, the order of the
elements matters. Catching a Pink salmon as your first fish brings you
to a whole new branch of the decision tree than drawing it after
catching another fish, and will therefore lead to different outcomes.

![A tree digram showing all the possible outcomes for a pond with a
Pink, Coho, and Chum
salmon"](media/diagrams/tree_diagram2.png){fig-alt="A tree diagram showing that there are sie possible outcomes from a pond with this combination of fish"}

When order is important to the arrangement of elements, it is called a
**permutation**. When you're using *all* the possible elements at a
time, the number of permutations in a set of $n$ elements is equal to
$n!$. We can see this in the above decision tree. If we were to only
have the option of catching *Pink*, *Coho*, or *Chum* salmon ($n = 3$),
then our number of permutations would be: $$
3! = 3 \times 2 \times 1 = 6 
$$

Now, what if we're calculating how many permutations can be made for
just a *portion* of the available elements? For example, how many
permutations of two different salmon can we create from a list of all
five salmon?

Just like before when we were calculating all the permutations for
catching five fish, at the first level, all five options are available.
Then there are four at the next level. $$
 \text{Permutations (P)} = 5 \times 4 = 20
$$

We stop after this because we are only looking for combinations of two.

**Question 9:** How many permutations of *three* different salmon can we
create from a list of all five salmon?

```{r}
#| tags: [question]
#| eval: false

answer_9 <- # your answer here

test_9()
```

```{r}
#| tags: [answer]
#| eval: false

answer_9 <- 60 # your answer here

test_9()
```

The general rule for counting permutations like this is the below
equation: $$
_{n}P_{r} = \frac{n!}{(n-r)!}
$$

where $n$ is the number of possible different elements (for example,
five possible species of salmon) and $r$ is the number of elements you'd
like in the permutation (for example, permutations of three different
salmon). $$
_{5}P_{3} = \frac{5!}{(5-3)!} = \frac{5!}{2!} = \frac{5\times4\times3\times2!}{2!} = 5\times4\times3
$$

> *Note*: This equation still holds true when the number of total
> elements is the same as the number of elements in our permutation.
> $0!$ is equal to 1, not 0! When $n = r$, the bottom of the equation
> becomes one, so our number of permutations is just equal to $n!$, as
> we determined in previous examples.

In calculating permutations, we assume we will not repeat elements.
There is only one Pink salmon in the pond, therefore it can only show up
once in our equation. The equation for permutations changes, however,
when the number the list of elements contains repetitions.

![A comparison of two ponds, displaying how the number of permutations
changes when there are repetitions in the
dataset"](media/diagrams/permutations_repetitions.png){fig-alt="A visualisation showing how the number of permutations was is reduced when a repeat of the element (two of the same salmon species) are introduced"}

Above are two fish ponds containing different fish, and the associated
permutations of three using all the fish in the pond. On the left, there
are no repeated species, but on the right, there are two pink salmon.
Using our equation without accounting for the repeated species, we would
see the same number of permutations for both ponds. Visually, though, we
can tell that some of the permutations on the left side have repeated.

To account for repetitions within a list of elements, we use the
following equation:

$$
P = \frac{n!}{a!\times b!\times c!}
$$

where $a$, $b$, and $c$ are the number of repetitions for different
elements. In the case of pond 2, we have just one repeating element
($a$) with two repetitions. If our list had other elements which have
repetitions, they would be included as $b$, $c$, and so on. So we would
calculate this as:

$$
P = \frac{n!}{a!}= \frac{3!}{2!} = \frac{3\times2\times1}{2\times1} = 3
$$

**Question d:** How many permutations could we generate from the
following list of elements: *Chum*, *Pink*, *Pink*, *Coho*, *Coho*,
*Sockeye*?

```{r}
#| tags: [question]
#| eval: false

answer_10 <- # your answer here

test_10()
```

```{r}
#| tags: [answer]
#| eval: false

answer_10 <- 180 # your answer here

test_10()
```

Through all of these examples, we have been assuming that the order of
the species we catch matters. This is explicit to permutations. When
determining the number of permutations, catching a Pink salmon and then
a Coho salmon is a different outcome than catching a Coho salmon and
then a Pink salmon. What about instances where order doesn't matter? The
selection of a set of elements when order is *not* important is referred
to as a **combination**.

Let's think back to our original pond, where we have one of each salmon
species, for a total of five salmon. When we calculated all of the
possible *permutations* for the order in which we catch the salmon, we
had a very high number. How would this number change when we are trying
to calculate the possible *combinations*? If you have exactly five
different fish, how many total combinations of five fish can you think
of?

You can probably see that although the number of permutations is very
high, in this case there is *only one* combination of fish species. This
is because the number of combinations from a list of elements can be
expressed as:

$$
_{n}C_{r} = \frac{_{n}P_{r}}{r!} = \frac{n!}{(n-r)! r!} 
$$

where $C$ is the number of combinations, $n$ is the number of possible
different elements, and $r$ is the number of elements which are being
taken at a time in the combination or permutation.

**Question 11:** Earlier we calculated the number of permutations of
three different salmon we could make from a list of all five salmon. How
many *combinations* of three could we make from a list of all five
salmon?

```{r}
#| tags: [question]
#| eval: false

answer_11 <- # your answer here

test_11()
```

```{r}
#| tags: [answer]
#| eval: false

answer_e <- 10 # your answer here

test_e()
```

## PART 3: DATA

Now that we have an understanding of counting and probability, we can
begin diving into one of the most important elements of good statistics:
data.

**Material Covered:**

-    Key components of data

-    Variables

-   Types of data

-    Basic data visualization

-    Describing data through central tendency

As our modern world continues to push forward technologically through
innovations in AI, modeling, algorithms, and personalized digital
advertising, you may be hearing lots about data in the world around you
at the moment. But what are data? Very broadly, data is information that
has been collected for some purpose or another.

Below are some helpful definitions to understand data:

-   **Dataset**: a structured collection of numbers and characters.
-   **Variable**: a characteristic, number, or quantity that can be
    measured.
-   **Observation**: all of the measurements for a given entity.
-   **Value**: a single measurement of a single variable for a given
    entity.

Data can be structured into datasets in many ways, but most commonly
will be presented as **tabular data**.

```{r}
head(salmon_data) # run this cell to view some tabular data!
```

In the above example, the rows represent observations. In this case,
each fish caught represents a unique observation. The columns,
meanwhile, are the variables in this dataset. Each observation has a
value assigned in each variable. The variables are what hold information
about the same characteristic (species, size, etc.) for all the
observations.

### Variable Types:

The type of characteristic and the way that it is recorded will change
the way that we can use a variable within a dataset. For example,
suppose we have a survey which asks participants the following questions
and store the answers in two separate variables.

> 1.  Are you concerned about the state of salmon stocks in BC (yes or
>     no)
> 2.  How much salmon do you consume in a year \_\_\_\_\_\_\_\_\_
>     *(grams)*

To report the data values you collected for question 1, it may make
sense '#/% of respondents are concerned about the state of salmon stocks
in BC', but would the same approach work for the second question?

Similarly, you may be able to calculate an average number of grams of
salmon consumed per person based on the data collected from the second
question, but no such calculation could be done for the data from the
first question.

This is because the data collected from these two questions are
*fundamentally different types of data*, and therefore have
fundamentally different ways of being reported, used, and interpreted.

Variables which contain information on what group or category an
individual (observation) belongs to are referred to as **qualitative
variables**, or sometimes categorical or nominal variables. Qualitative
data generally answer descriptive questions (what, where, how) and often
take the form of text responses - like question 1 in our example.

By nature of being qualitative, these types of data are more limited in
their applications, but be aware: just because a qualitative variable is
more limited in its application does not necessarily make it less useful
or informative.

**Quantitative variables**, meanwhile, contain numerical data which
express a measurable quality of the observations. In addition to the
numeric values, these variables are described using a form of
measurement units ---for example, centimeters, kilometers per hour, or
degrees Celsius--- which provide meaning to the numeric data by
indicating the scale of measurement.

Within quantitative variables, there are important distinctions. Think
back to our probability model data, where we explored the mean expected
value for the number of salmon we would catch in a day. The data from
which this probability model was drawn would have recorded a numeric
value, with *salmon* as the measurement unit, and is therefore
quantitative data.

However, there is a limitation on the values that this response could
take on in that it is only possible to catch a whole fish at a time.
This quantitative variable is therefore a **discrete variable**, as we
can list all possible outcomes. Conversely, the variable that would have
recorded the length of the fish was not limited in this way. A salmon
could be any fraction of a centimeter in length, and the number of
values that the variable could take on is limited only by the precision
of our measuring tools. This type of quantitative variable is called a
**continuous variable**.

Keep in mind that just because a variable is recorded as a number does
not necessarily mean it is a quantitative variable. In some cases,
qualitative variables may be labeled using numerals, or converted to
numbers for ease of management, but they are still at their core
recording categories of the observations.

In other cases, qualitative data is collected using a scale. For
example, a survey may ask you to rate an experience on a scale from one
to ten. Although the value assigned is a numeric response, it is still
qualitative data, as this scale is a unitless creation, which has no set
distances between values. The responses on the scale cannot therefore be
interpreted numerically, but is still useful as a way of categorizing
the way that participants felt about an experience. This particular type
of variable is referred to as an **ordinal variable**.

Below is a list of sample variables that could be collected.

|     | Variable                                                                       |
|---------------------|---------------------------------------------------|
| 1   | driving speed, in kilometers/hour                                              |
| 2   | the recorded hair color of an individual                                       |
| 3   | the number of bedrooms in a house                                              |
| 4   | an individual's perceived understanding of a topic, rated on a scale from 1-10 |
| 5   | the banking institution of an individual                                       |
| 6   | amount of fertilizer applied to a field, measured in kilograms per hectare     |
| 7   | number of clients served at a business in a day                                |
| 8   | numerical area code of a surveyed group                                        |

**Question #12:** Which of the above variables are continuous
quantitative variables. Please list all applicable, separated by commas,
in ascending order

```{r}
#| tags: [question]
#| eval: false

answer_12 <- '' #your answer here ('1,2,3..')

test_12 ()
```

```{r}
#| tags: [answer]
#| eval: false

answer_12 <- '1,6' #your answer here ('1,2,3..')

test_12 ()
```

**Question 13:** Which of the above variables are qualitative variables?
Please list all applicable, separated by commas\*\*

```{r}
#| tags: [question]
#| eval: false

answer_13 <- '...' #your answer here ('1,2,3..')

test_13()

```

```{r}
#| tags: [answer]
#| eval: false

answer_13 <- '2,4,5,8' #your answer here ('1,2,3..')

test_13()

```

### Data Types

There are many different types of data, which depend on the kinds of
values stored in the variables.

**Cross-sectional data** are data which present information on multiple
observations at a single moment in time.

![A table of sample cross sectional
data.](media/diagrams/cross_sectional_data.png){fig-alt="Sample data recording information about participants' age, height, city of residence, etc."}

**Time series data** are data which contain values for a single
individual over time. For example, the below table shows the changes
over time to a single salmon stock.

```{r}
head(population_data)
```

**Panel data** are data which contain values for multiple observations
over time.

![A table of sample panel
data.](media/diagrams/panel_data.png){fig-alt="Sample data recording information about voting turnout and population for two different cities, for multiple years."}

> *Think Deeper*: There are lots of data being collected, made
> available, and strategically presented from a variety of different
> sources. When interpreting data, you'll need to be exercising your
> information literacy skills, and making sure that you're critically
> examining the source of the dataset. You may want to ask yourself
> questions like: who created this dataset, what interest do they have
> in this information, and how was this information collected?

### Data Visualizations

One common way of beginning to understand the data is through data
visualization. Good data visualizations should always answer a
well-thought-out and relevant research question. Good data visualization
should also exist as a standalone explanation - that is, it should
provide readers with a clear understanding of both the research question
at hand and its answer in a way that doesn't require further
explanation.

|     | Quality                                                                                     |
|---------------------|---------------------------------------------------|
| 1   | use design features to manipulate readers into believing a particular narrative of the data |
| 2   | remember that a visualization's accuracy is only as good as the data is                     |
| 3   | use data visualization to tell the story of the data truthfully                             |
| 4   | label your axes in font sizes that are readable and use descriptive titles                  |
| 5   | choose colours that are very similar each other when trying to distinguish 2 variables      |

**Question 14:** From the above list of qualities, which do you think
you should keep in mind as good practices when creating visualizations?
Please list selected answers in ascending order, separated by commas.

```{r}
#| tags: [question]
#| eval: false

answer_14 <- '...' # your answer here ('1,2,3..')

test_14()
```

```{r}
#| tags: [answer]
#| eval: false

answer_14 <- '2,3,4' # your answer here (1,2,3..)

test_14()
```

There are many different kinds of visualizations, but we'll just focus
on some of the most common types.

> A **scatter plot** visualizes the relationship between two
> quantitative variables

This plot works great when we are interested in showing relationships
and groupings among variables from relatively large datasets

```{r}
scatter <- ggplot(data = salmon_data, aes(x = length, y = mass, color = species))+
  labs(x ="length of fish (cm)",
       y = "mass of fish (kg)", 
       title = "Mass and Length of Salmon, by Species")+
  scale_x_continuous() +
    theme (
    plot.title = element_text(size = 24),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)) +
  theme(aspect.ratio = 1.5
   )
  
scatter + geom_point(pch = 19, cex = 2.6) + scale_color_brewer(palette="Set2")

```

> A **line plot** visualizes trends with respect to an independent,
> ordered quantity (e.g., time).

This plot works great when one of our variables is ordinal (time-like)
or when we want to display multiple series on a common timeline.

```{r}
line <- ggplot(data = population_data, aes(x = year, y = run_size, color = region))+
  labs(x ="Year",
      y = "Run Size (conservation units)",
      title = "Chinook Salmon Yearly Run Size, by Region") +
  theme (
    plot.title = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 12)) 

  
line + geom_line(size = 1.5) + scale_color_brewer(palette="Accent")
```

> A **bar plot** visualizes comparisons of amounts.

This plot works great when we are interested in comparing a few
categories as parts of a whole, or across time.

```{r}
bar <- ggplot(data = salmon_data, aes(species, fill = species))+
  labs(x ="Salmon Species",
       y = "Count",
       title = "Count of Salmon by Species") +
  theme (
    plot.title = element_text(size = 24),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 16)
    ) 

bar + geom_bar() + scale_fill_manual(values = c("#EAB676","#CADFBB","#D0E5E9","#F8DDF2","#FFA095"))  
```

> A **histogram** visualizes the distribution of one quantitative
> variable

This plot works great when we are working with a discrete variable and
are interested in visualizing all its possible values and how often they
occur. They can also be used for continuous variables as in the below
graph.

```{r}
histogram <- ggplot(data = salmon_data, aes(length)) +
  labs(x ="Salmon Length (cm)",
       y = "Count",
       title = "Length of Salmon") +
  theme (
    plot.title = element_text(size = 24),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 16)
    ) 

histogram + geom_histogram(fill = "darkseagreen3", color = "darkseagreen4", binwidth = 5)
```

### Summary Statistics and Central Tendency

Importantly, data also allows us to explore **statistics**, which is the
field of mathematics that is concerned with the analysis and
manipulation of data. There are a number of different analyses within
statistics that each play a role in interpreting data and driving
decisions. One common way of using data is the computation of **summary
statistics**, which are statistics that describe the general behaviour
of variables.

#### Central Tendency - Mean

In statistics, central tendency refers to the idea of how different
interpretations of the term "middle" can be used to describe a
probability distribution or dataset. In this notebook, we'll think about
central tendency in terms of numerical values which describe a given
subset of the dataset. This concept is important because we often deal
with incredibly large datasets that are too big to describe in their
entirety.

The first, and most referenced measure of central tendency is the
**sample mean** (also referred to as the **arithmetic mean**). The mean
of a variable is the average value of that variable, which can be found
by summing together all values that a variable takes on in a set of
observations and dividing by the total number of observations used. This
is an intuitive measure of central tendency that many of us think of
when we are trying to describe data. The formula for the sample mean is
below.

$$ 
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i =  \frac{\text{Sum of All Data Points}}{\text{Total Number of Data Points}}
$$

```{r}
view(salmon_data) #run this cell to call the table of data from before  
```

**Question 15:** What is the mean length of the first 10 salmon in this
dataset? Please round your answer to two decimal places.

```{r}
#| tags: [question]
#| eval: false

answer_15 <- # your answer here 

test_15()
```

```{r}
#| tags: [answer]
#| eval: false

answer_15 <- 63.4 # your answer here 

test_15()
```

#### Central Tendency - Median

Another common measure of central tendency is the **median**. The median
is the value which exactly splits the observations for a variable in our
data set in half when ordered in increasing (or decreasing) order. For
instance, if we had observations for `variable 1` of 60000, 45000, and
72000, our median value for `variable 1` would be 60000 because there is
exactly one observation above (70000) and one observation below (45000)
this value.

To find the median of a variable, we order all of our observations in
ascending (or descending) order, then find the value in the middle of
this ordered list. This works great when we have an odd number of
numbers to list (i.e. 9 observations). However, for a list which is of
even length (i.e. has 10 observations), we must take the middle *two*
observations and take their arithmetic mean (sound familiar!). A formula
for finding the median in each of these two cases is below.

Arrange data in ascending (or descending) order. Let $n$ be the number
of data points. If $n$ is odd, then:

$$
\text{Median} = \frac{n+1}{2}\text{th data point}
$$

Otherwise,

$$
\text{Median}  = \frac{1}{2} [\frac{n}{2}\text{th data point} + (\frac{n}{2} + 1)\text{th data point}]
$$

In other words, the median value is the value that is, quite literally,
in the middle of the dataset. Half of the values in the dataset are
above it and half are below it.

The median alone, however, is limited in its use. Take for example the
two datasets below.

| Obs | variable 1 | variable 2 |
|-----|------------|------------|
| 1   | 1.1        | 1.8        |
| 2   | 0.8        | 1.8        |
| 3   | 1.6        | 1.7        |
| 4   | 2.1        | 1.7        |
| 5   | 2          | 1.5        |
| 6   | 2.2        | 2.6        |
| 7   | 1.7        | 2.4        |
| 8   | 1.8        | 3.3        |
| 9   | 2.1        | 2.3        |

**Question 16:** What is the median value for `variable 1`? Please give
your answer to one decimal place.

```{r}
#| tags: [question]
#| eval: false

answer_16 <-  # your answer here 

test_16()
```

```{r}
#| tags: [answer]
#| eval: false

answer_16 <- 1.8 # your answer here 

test_16()
```

**Question 17** What is the median value for `variable 2`?

```{r}
#| tags: [question]
#| eval: false

answer_17 <- # your answer here 

test_17()
```

```{r}
#| tags: [answer]
#| eval: false

answer_17 <- 1.8 # your answer here 

test_17()
```

What do you notice about these two values?

To help understand why the median can be an incomplete description of
the data by itself, let's see what these two variables look like as
histograms.

```{r}
 #run this cell to see histograms!

par(mfrow=c(1,2))   
hist(variable_1,                   
     breaks= 5,                
     xlab = "Variable 1", 
     main="Histogram Variable 1",                    
     col = "darkseagreen3",  
     xlim = c(0.5, 3.5) )    
hist(variable_2,                  
     breaks= 5,                
     xlab = "Variable 2", 
     main="Histogram Variable 2",                    
     col = "darkslategray3",
     xlim = c(0.5, 3.5) )  
  
```

As we can see from these graphs, these data are very distributed very
differently!

> *Think deeper*: Can you calculate the means for these data? How does
> knowing both values of central tendency change how you understand the
> distribution of the data?

## PART 4: INFERENTIAL STATISTICS

**Material covered:**

-   Regression models

-   Samples vs. population

-   Point estimates and statistical inference

-    Sampling variability and normal distribution

### Regressions

Up until this point, all of these techniques have looked at variables in
isolation of one another. In reality though, the variables within
datasets are often connected to each other through relationships with
varied complexity. In statistics, **regressions** present a way to
analyze these relationships.

Let's return to our fishing dataset. In BC, there are rules in place
which limit the size of salmon that recreational fishers are permitted
to take home. The size limits are based on a salmon's length in
centimeters measured from the tip of the nose to the tail fork. These
regulations have been introduced to help ensure that an adequate number
of mature salmon are able to return to their spawning grounds. For
Chinook salmon in the Lower Mainland region of BC, this means a maximum
of ten salmon over 50 cm in length.

If you are catching salmon for food, however, it might not be
immediately clear what kind of limitation this places on how much salmon
you have available to eat; someone eating salmon may not typically think
about it in terms of length. They instead might want to think about this
regulation as the *relationship* between the salmon's length and its
weight.

To begin exploring some of this relationships, let's start off with a
visualization:

```{r}
plot(salmon_data$length,
     salmon_data$mass, 
     xlab = 'length (cm)', # provide a descriptive label
     ylab = 'mass (kg)', # provide a descriptive label
     main = 'Scatter plot of salmon length vs. mass', # provide a descriptive title
     col = 'darkorchid4'
     ) 
```

You can probably tell that there is definitely some relationship between
length of the salmon and its mass- but it can be difficult to understand
completely using a scatter plot alone. There are far too many points,
and too much going on.

### Regression Models

This is where a *regression model* comes in. A regression model
specifies the relationship between two variables. For example, a linear
relationship would be:

$$
M_i = \beta_0 + \beta_1 W_i
$$

Where $M_i$ is the market income of individual $i$, and $W_i$ is their
wage. We normally refer to $M_i$ as the **outcome** variable, and $W_i$
as the **explanatory** variable; these may also be referred to as
*dependent* and *independent* variables, depending on the field.

A model like this is our description of what this relationship is - but
it depends on two unknowns: $\beta_0$, $\beta_1$.

-   The $\beta_0$ and $\beta_1$ are **parameters** of the model: they
    are numbers that determine the relationship (intercept and slope)
    between $M_i$ and $W_i$. In our model $\beta_0$ is the intercept,
    which tells us where the line hits the y-axis. $\beta_1$ is the
    slope, which tells us how rapidly \$M_i \$ changes with respect to
    $W_i$
-   This is a linear relationship because the model we have specified is
    linear in the coefficients, giving us an equation of a straight line
    through the data.

Let's try this out on our own dataset. Use the below sliders to move the
line around. See if you can get a line running through all the data
points.

```{r}
plot(salmon_data$length,
     salmon_data$mass, 
     xlab = 'length (cm)', # provide a descriptive label
     ylab = 'mass (kg)', # provide a descriptive label
     main = 'Scatter plot of salmon length vs. mass', # provide a descriptive title
     col = 'darkorchid4'
     ) 

abline(a,b, col = "black")
```

```{r}
a = -1
b =  -0.4
 #Change these numbers and re-run the cell!

```

Of course, we will never get a line which goes through every point; it
is unlikely, if not impossible, for ($M_i = \beta_0 + \beta_1 W_i$) to
completely explain everything about our data. We need to include a term
which captures everything which is *not* described by this relationship.
This is called the *residual* term (meaning "leftover"). The
$\epsilon_i$ is the **residual**: a component that corresponds to the
part of the data which is *not* described by the model

Conceptually, we can think about a regression as two parts: the part of
the relationship explained by our model ($M_i = \beta_0 + \beta_1 W_i$)
and the part which is not explained ($\epsilon_i$). Even though not
everything is explained by the model, there is still a "best" line that
we can draw through the data. In fact, the residuals help in being able
to know what it is. In simplified terms, the residuals assists in
mathematically determining the line of best fit by providing a measure
of minimizing the distance between all the individual data points and
the line. We'll do this automatically and fit it to the plot below.

```{r}

plot(salmon_data$length,
     salmon_data$mass, 
     xlab = 'length (cm)', # provide a descriptive label
     ylab = 'mass (kg)', # provide a descriptive label
     main = 'Scatter plot of salmon length vs. mass', # provide a descriptive title
     col = 'darkorchid4'
     ) 

abline(slr, col = "black")
text(70, 1, 'y = -3.7192 + 0.1036x')

```

We write the complete *regression equation* by combining the two parts
of the model:

$$
M_i = \beta_0 + \beta_1 W_i + \epsilon_i
$$

Keep in mind though that this equation represents the "true"
relationship that exists. Meanwhile, when we create models, we are
*predicting* the values as best we can based on our data. Therefore, we
would write are linear models in the following format:

$$
\hat{y} = b_0 + b_1x
$$

In this model, the $b_0$ and $b_1$ are no longer parameters, but
**coefficients** of the linear model. You'll also notice that $M_i$ has
changed to $\hat{y}$ (called 'y-hat'), indicating that it is the
**predicted value** of the model based on value $x$.

The goal of regression analysis is to:

1.  Estimate this equation (and especially the model parameters) as
    accurately as possible.
2.  Learn about the relationship between $M_i$ and $W_i$ from the
    results of that estimation.

**Question 18:** What is the estimated value of the slope in our linear
regression? Please give your answer to 4 decimal places.

```{r}
#| tags: [question]
#| eval: false
answer_18 <- # your answer here

test_18()
```

```{r}
#| tags: [answer]
#| eval: false
answer_18 <- 0.1036 # your answer here

test_18()
```

**Question 19:** Using the model we have developed, predict how many
grams of fish we can expect from a salmon measuring 52.54 cm in length.
Please round your answer to two decimal places.

```{r}
#| tags: [question]
#| eval: false
answer_19 <- # your answer here

test_19()
```

```{r}
#| tags: [answer]
#| eval: false
answer_19 <- 1.72 # your answer here

test_19()
```

### The Basis of Statistical Inference

Now that we have an understanding of a simple regression, you may have
realized that we have begun to go beyond simply what we can learn from
the data about the dataset itself, but what we can *infer* from the data
about information outside of the dataset. In this last section, we will
start to unpack the basis of statistical inference, and how it's
possible.

One of the common uses for data is to help to understand some 'truth'
about a particular group. See the following example of a headline from
the [Pacific Salmon
Foundation](https://psf.ca/news-media/86-bc-residents-show-high-levels-concern-about-declining-salmon-stocks/):
"86% of B.C. residents show high levels of concern about declining
salmon stocks.

Think about this statement for a minute, and ask yourself: how do we
know this? Did the Pacific Salmon Foundation ask each and every BC
resident what they thought about the declining salmon stocks? The answer
of course is no, they did not- how could they? But that doesn't mean
that this statement is incorrect. They've just used statistics to arrive
at this conclusion. When it's not practical to measure this information
directly, *statistical inference* gives us a method of evaluating it
from a subset of the population.

For example, we might want to know how many BC residents are concerned
about the state of the province's salmon stocks. We are interested in
making a conclusion about *all* BC residents, which we would call the
**population**. In statistics, population refers to the complete group
of individuals or instances we are interested in studying. From this
population, we would like to compute a quantity, which is the proportion
of individuals who are concerned about the salmon stocks in BC. This
value we are trying to compute is referred to as a population
**parameter**. A parameter is a numerical characteristic which is true
of the *entire* population.

As we've established, a parameter may be challenging or impossible to
compute directly. We don't have the resources to ask every BC resident
to answer this question. Instead, we can estimate this information by
taking the measurements from a representative **sample** of the
population. A sample is a subset of individuals which have been selected
from the population. When we use a sample to summarize a characteristic
of the data, it is called a **statistic**. We understand that this
sample can be used to make conclusions about the larger population, and
this is the basis of **statistical inference**.

![A diagram which shows that the sample is a representation of the
population.](media/diagrams/population_sample.png){fig-alt="A diagram displaying how sample data is taken from a larger populuation"}

The above statistic is given as a **proportion**, and expresses the
share of an outcome of a variable in relation to the whole. We formally
use $\pi$ to denote a proportion when we're talking about a proportion
of the population, and $\hat{p}$ when we're talking about the proportion
of a sample. We can calculate proportions using:

$$
\hat{p} = \frac{x}{n}
$$

Where $n$ is the number of 'successes' for the proportion of interest
(in our case, how many people said that they *are* concerned about BC's
salmon stocks) and $x$ is the total number of observations in the
dataset (how many people were surveyed).

> *Note*: When we talk, we would normally describe proportions using
> percentages. People frequently say things like "we've completed 60% of
> our tasks", or "12% of our clients were unsatisfied with their
> service", as this helps when communicating mathematical information
> verbally. When writing things formally in statistics, however, we use
> decimal points to convey the same information. For example, 60%
> becomes 0.6, and 12% becomes 0.12.

Suppose we want to better understand the reasons that the entire
population of BC residents harbours concerns for salmon stocks. To
determine this, we poll 100 randomly selected British Columbians and
arrive at the following dataset:

![A table of sample data of the primary reason the polled participants
feel concerned about salmon
stocks.](media/diagrams/sample_1_data.png){fig-alt="Not concerned = 13, ecosystem/environmental = 44, livelihood = 15, consumption = 28, total = 100"}

**Question 20:** What is the proportion of this sample who are concerned
about salmon stocks for reasons related to ecosystem health? Please
provide your answer as a decimal value (e.g. 50% = 0.5).

```{r}
#| tags: [question]
#| eval: false
answer_20 <- # your answer here

test_20()
```

```{r}
#| tags: [answer]
#| eval: false
answer_20 <- 0.44 # your answer here

test_20()
```

Because we used a sample to arrive at this number and not the entire
population, this calculation is only a **point estimate** of the
population parameter, meaning it's out best guess at a single value. Say
we repeat this poll on a new group of British Columbians, and obtain the
following dataset:

![Another table of sample data of the primary reason the polled
participants feel concerned about salmon
stocks.](media/diagrams/sample_2_data.png){fig-alt="For the polled participants, not concerned = 17, ecosystem/environmental = 47, livelihood = 23, consumption = 13, total = 100"}

**Question 21:** What is the proportion of *this* sample who are
concerned about salmon stocks for reasons related to ecosystem health?
Please provide your answer as a decimal value (e.g. 50% = 0.5).

```{r}
#| tags: [question]
#| eval: false
answer_21 <- # your answer here

test_21()
```

```{r}
#| tags: [answer]
#| eval: false
answer_21 <- 0.47 # your answer here

test_21()
```

As you probably expected, these point estimates are not the same due to
**sampling variability**, or sample-to-sample differences. This kind of
variability is both common and expected when collecting data.

So, if there's variability in our point estimates, how can we ever know
the population parameter? In truth, there will always be uncertainty in
statistics, and we can only ever use estimates; however, statistical
analysis provides us with methods to make these estimates robust.

Let's take another three samples and see what we have.

| Sample | Proportion |
|--------|------------|
| 1      | 0.44       |
| 2      | 0.47       |
| 3      | 0.47       |
| 4      | 0.39       |
| 5      | 0.45       |

We can display these results as a histogram to better understand them.

```{r}
histogram_5_samples #run this cell to view the our five_sample_histogram
```

Suppose we kept repeated this with new samples asking this same
question, and continued calculating this proportion. The results are
displayed below.

```{r}
#| eval: false
View(sample_data) ## run this cell to view the dataset
```

Now change the histogram to add more lines of data to the histogram in
increments of ten in the indicated place below ("change me"), until
you've included all 100 observations. Re-run the whole cell after every
addition to see how adding more samples changes the histogram.

```{r}
#| eval: false
histogram_100_samples <- ggplot(data = sample_data[1:100 #(change me --> 1:10, 1:20, 1:30... etc.)!
                                           
                                           ,],  aes(x = Proportion)) +
  geom_histogram(fill = "dodgerblue3", color = "lightgrey", bins = 8) +
  labs(x = "Sample proportions", y = "Count", main = "Histogram: 100 Samples") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 22))

histogram_100_samples
```

Notice how the shape of the histogram changes? With this larger set of
data, it looks far more symmetrical, with a single peak close to 0.5.
This bell-curve shaped is called a **normal distribution** of data.

If we were to continue polling until we included all possible samples of
a given size ($n$), we would see it become even more symmetrical, and
centered perfectly on $\pi$, the true proportion of BC residents who are
concerned about salmon stocks for reasons primarily related to
environmental or ecosystem health. This is a mathematically proven
concept which is fundamental to inferential statistics.

Knowing that this is the case does not necessarily mean that the
proportion we calculate from our sample will always be very close to the
true parameter of the population. Sampled values have the potential to
actually be quite different from their corresponding parameters.

![An example of a sampled value which is very different from the
majority of other
samples](media/diagrams/histogram_100.jpg){fig-alt="A bin in the histogram records a value much higher than most of the other values"}

Nevertheless, knowing how these values are distributed around the
population parameter allows us to make strong conclusions about how
accurate our statistics are. Our sample statistic could be very
different from the true value of $\sigma$. But is it likely? How likely?
Conversely, what is the likelihood that it is very close? With enough
information about our sample, these questions can be answered through
calculation!

This is because statistics is directly connected to probability.
Statistical inference is based on understanding the *probability* that
with any given sample size, ($n$), the true characteristic of the
population ($\sigma$ or $\mu$) will fall a certain distance from the
sampled value ($p$ or $\bar{p}$).

## Conclusion

As you can see, the world of probability and statistics is incredibly
fascinating, with so many important applications. In this notebook, we
learned about probability, counting, data, and statistics using the
state of BC salmon as a case study to provide insight into how these
skills can be used to understand natural resource management and
ecological conservation efforts. The importance of salmon to many people
who call our province home means that it is critical to continue to use
data and statistics to thoughtfully address the threats and
opportunities for the future of salmon in BC.

After having completed this introductory course, you should have a solid
foundation of knowledge about the relationships between probability,
statistics, and data, and some of the applications of these concepts in
the world around us.
